Does It Scale? Static Output Approximation of PHP Web Applications Using Symbolic Execution

### 2017-01-13 paper.tex
2.2 refactored

% The default list of authors is too long for headers}
\renewcommand{\shortauthors}{S. Mühlbauer et al.}

@@

\subsection{Static Output Approximation for PHP} \label{sec:workflow}
Symbolic execution is an analysis technique not dedicated to a particular
programming language. For dynamic web applications, in our case for web
applications written in PHP, it can be used to approximate all possible client
page output. Before we present in detail existing tools that aim to ease
development and maintenance of PHP web applications, we illustrate how output
approximation using symbolic execution is used by starting with an example.

@@

Figure \ref{fig:code_example} shows two code listings. The snippet of server-side web
application code in Figure \ref{fig:web_application} contains both HTML and PHP. First, the
function \texttt{make\_titles} defined in lines 7 – 13 prints the string
\texttt{\$arg} passed as an argument between zero and an arbitrary number of
times. Second, the function is indirectly called using the built-in function
\texttt{call\_user\_func} in line 15 with the string \texttt{Headline} and
\texttt{4} as arguments.  Finally, the input from the form in lines 1 – 4 is
used: The variable \texttt{\$greeting} consists of three concatenated strings,
\texttt{Hello} the value of input field \texttt{name}, and an exclamation mark.
For this server-side code snippet there are two possible client page output
variants: If the input field \texttt{name} is empty at runtime, line 19 is
executed and \texttt{No name entered!} is printed, or, if a name was entered,
line 21 is executed and the greeting constructed in line 17 is printed.

%C. Introduction to symbolic execution of PHP code
The following symbolic execution semantics for PHP features  based on \cite{Nguyen:2014:BCG:2635868.2635928}
illustrate the necessary modifications made to handle the extent of ambiguity
that comes with handling symbolic values. In addition to normal output, the
output approximation contains preprocessor directives to express
repeatable output as well as output variants. All approximated client page
output corresponding to Figure \ref{fig:web_application} consisting of two different variants
is illustrated in Figure \ref{fig:output_example}.
First, loops in the previous execution semantics are only executed once instead
of an arbitrary number of times. Since a loop condition can contain symbolic
values, it might not be feasible to determine when a loop terminates. Hence,
any output constructed during the single loop iteration is highlighted
(\texttt{repeat \ldots  endrepeat}) to possibly be repeated an arbitrary number
of times. The output in line 7 is marked to be repeated an arbitrary number of times, as the number
of loop iterations cannot be determined statically. Second, a similar procedure
applies to recursive functions. Once called, a recursive function is  also only
executed to recursion depth one per call, and any subsequent invocation of that
function returns a symbolic value. This, again, is due to the difficulty to
determine the exact recursion depth similar to the number of loop iterations.
Third, since the symbolic execution engine applies a static analysis, any form
of user interaction providing input data as well as any form of input referred
to from the deployment context of the web application, i.e. configuration files
or databases, is substituted by a symbolic value. The context variable
\texttt{\$\_POST} represents input submitted to the server and, hence, is
symbolic as every program input is per definition substituted by symbolic values. Finally, the
output in lines 11 and 13 respectively are marked as alternatives
(\texttt{if \ldots else \ldots endif}), as symbolic execution of lines 18 – 22
in Figure \ref{fig:web_application} will explore both different branches.

>>>>

% The default list of authors is too long for headers}
\renewcommand{\shortauthors}{\textcolor{blue}{put shortauthors here}}

@@

\subsection{Static Output Approximation for PHP} \label{sec:workflow}
Symbolic execution is not dedicated to a particular
programming language. For dynamic web applications, in our case for web
applications written in PHP, it can be used to approximate all possible client
page output variants. Before we present in detail existing tools that aim to
ease development and maintenance of PHP web applications, we illustrate how output
approximation using symbolic execution is pursued by starting with a small
example.

@@

Figure \ref{fig:code_example} shows two code listings. The snippet of server-side web
application code in Figure \ref{fig:web_application} contains both HTML code
(input form in lines 1 -- 4) as well as PHP code: First, the function
\texttt{make\_titles} defined in lines 7 – 13 prints the string \texttt{\$arg} passed as an argument between zero and an arbitrary number of
times. Second, the function is indirectly called using the built-in mechnism
\texttt{call\_user\_func} in line 15 with the string \texttt{Headline} and
\texttt{4} as arguments.  Finally, the input from the form in lines 1 – 4 is
used: The variable \texttt{\$greeting} consists of three concatenated strings,
\texttt{Hello}, the value of input field \texttt{name}, and an exclamation mark.
For the greeting there are two possible client page output
variants: If the input field \texttt{name} is empty at runtime, line 19 is
executed and \texttt{No name entered!} is printed, or, if a name was entered,
line 21 is executed and the greeting constructed in line 17 is printed.

The following symbolic execution semantics for PHP features based on
previous work \cite{Nguyen:2014:BCG:2635868.2635928} illustrate the necessary
modifications and extensions made to handle the level of ambiguity that comes with handling symbolic values. In addition to normal output, the
output approximation contains preprocessor directives to express
repeatable parts as well as output variants. All approximated client page
output corresponding to Figure \ref{fig:web_application} consisting of two different variants
is illustrated in Figure \ref{fig:output_example}.
First, for loops in the previous execution semantics
\cite{Nguyen:2014:BCG:2635868.2635928} only one iteration is executed instead of
an arbitrary number of iterations. Since a loop condition can contain symbolic
values, it might not be feasible to determine whether, and if so, when a loop
execution terminates. Hence, any output constructed during the single loop
iteration is highlighted (\texttt{repeat \ldots endrepeat}) to possibly be repeated an arbitrary number of times. The output in line 7 is marked to be repeated an arbitrary number of times, as the number of loop iterations cannot be determined statically. Second, a similar procedure
applies to recursive functions. Once called, a recursive function is  also only
executed to recursion depth one per call, and any subsequent invocation of that
function returns a symbolic value. This, again, is due to the difficulty to
determine the exact recursion depth similar to the number of loop iterations.
Third, any form of user interaction providing input data as well as any form of input referred
to from the deployment context of the web application, i.e. configuration files
or databases, is assumed to be symbolic. The context variable
\texttt{\$\_POST} represents input submitted to the server and, hence, is
symbolic as every program input is per definition substituted by symbolic values. Finally, the
output in lines 11 and 13 respectively are marked as alternative variants
(\texttt{if \ldots else \ldots endif}) as symbolic execution of lines 18 – 22
in Figure \ref{fig:web_application} will explore both different branches.
